### 2017-10-02 chapters/introduction/mechanics.tex
Added paper 3 and update introduction
Refer to newly published paper about comparison of active stress and active strain
Incorporate Sams suggestions

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F(\Xvec) = \frac{\partial \varphi}{\partial \Xvec} = \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

@@

\subsection{Balance laws and transformations}
In this section we will cover some basic transformation used to
derive the fore-balance equations for the mechanics of the heart.

@@

\subsubsection{Conservation of linear momentum}
Newton's seconds law states that the change in linear momentum equals
the net impulse acting on it. For a continuum material with constant
mass density $\rho$ this implies that
\begin{align}
  \int_{\omega} \rho \dot{\mathbf{v}} \mathrm{d}v = \mathbf{f},
  && \mathbf{f} = \int_{\partial \omega} \mathbf{t} \mathrm{d}s
     + \int_{\omega} \mathbf{b} \mathrm{d}v,
     \label{eq:cons_lin_mom}
\end{align}
where $\mathbf{v}$ is the spatially velocity field, $\mathbf{t}$ is
the traction acting on the boundary, and $\mathbf{b}$ is the body
force. From \emph{Cauchy's stress theorem} we know that there exist a
second order tensor $\sigma$, known as as Cauchy stress tensor that is
related to the traction vector by $\mathbf{t} = \sigma \mathbf{n}$,
where $\mathbf{n}$ is the unit normal in the current configuration.
Using the divergence theorem we get.

\begin{align*}
  \int_{\partial \omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom}. We arrive at
Cauchys momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) are negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

\subsubsection{Conservation of angular momentum}
Just like linear momentum, the angular momentum is also a conserved
quantity. We will not go through the derivation of resulting
equations, but state that as a consequence, the Cauchy stress tensor is
symmetric
\begin{align}
  \sigma = \sigma^T.
\end{align}

@@

Even though experimental studies have indicated a visco-elastic behavior
of the myocardium \cite{dokos2002shear, gultekin2016orthotropic}, a
common assumption is to consider a quasi-static behavior, meaning that
the inertial term in \eqref{eq:chauch_momentum_eq} is negligible and
static equilibrium can be achieved at any time. Therefore
it is also possible to model the myocardium as a hyperelastic material
which is a type of elastic material.
This means that we postulate the existence of
a strain-energy density function $\Psi:\mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$, and that stress is given by the relation
\begin{align}
\FPK = \frac{\partial \Psi(\F)}{\partial \F}.
\end{align}
Since stress has unit Pa, we see that the strain-energy density
function is defined as energy per unit reference volume, and has units
$\frac{\text{Joule}}{m^3}$. 
The strain-energy density function relates  the amount of
energy that is stored within the material in response to a given
strain. Hence, the stresses in a hyperelastic material with a given
strain-energy density function, depends only on the strain, and not the
path for which the material deforms. On the contrary, if the model had
been visco-elastic we would expect to see hysteresis in the
stress/strain curve, but this is not possible for a hyperelastic
material.

\begin{remark}
  The second law of thermodynamics states that the total entropy
  production in a thermodynamic process can never be negative. Elastic
  materials defines a special class of materials in which the entropy
  production is zero. Within this thermodynamic framework the
  strain-energy density function coincides (up to a constant) with the
  Helmholtz free energy density.
\end{remark}

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.

@@

However, from a physical point of view this requirement is too strict
\cite{ball1976convexity}. A slightly weaker requirement is the strong
ellipticity conditions which states that \eqref{eq:response_posdev}
should hold for any $\mathbf{H}$ of rank-one, which is analogous to
say that the strain energy function is rank-one convex.

@@

The myocardium is composed by small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts, this
perfused blood is squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. Such materials are
referred to as compressible. When the 
volume are preserved we refer to the material as incompressible.
Since 2-4\% is very little, a common assumption in cardiac mechanical modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompressible. The reason for this choice is
purely numerical.

For an incompressible material, only isochoric motions are
possible. This means that volume of the material do not change during
any deformation and hence we have the constraint 
\begin{align}
  J = \det(\F) = 1.
  \label{eq:incompressible_cons}
\end{align}
The constraint \eqref{eq:incompressible_cons} can be imposed by
considering the modified strain energy function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + J p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = J^{-1} \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

@@

\subsubsection{Uncoupling of volumetric and isochoric response}
The total strain energy function in \eqref{eq:incomp_strain_energy}
can be written as a sum of isochoric and volumetric components. Let
\begin{align}
  \F =  \F_{\mathrm{vol}} \F_{\mathrm{iso}},
\end{align}
then $ \F_{\mathrm{vol}} =
J^{1/3}\I$ and $\F_{\mathrm{iso}} = J^{-1/3}\F$. For
compressible material (i.e with $J \neq 1$) it is important to consider
only deviatoric strains in the strain-energy density function, so that
$\Psi = \Psi_{\mathrm{iso}}(\F_{\mathrm{iso}}) +
\Psi_{\mathrm{vol}}(J)$. For incompressible material ($J = 1$), we
have $\F_{\mathrm{vol}} = \I$ so that such a decomposition seems
unnecessary. However, a similar decomposition has shown to be
numerically beneficial \cite{weiss1996finite}. Note that, in this case, a similar
decoupling of the stress tensors has to be done.

@@

Choosing the correct boundary conditions for the model is essential,
and the choice should mimic what is observed in the reality. To
physiologically constrain the ventricle in a correct way is difficult,
and different approaches has been proposed.
The boundary condition at the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure. For the left ventricle we have
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lVendoCur, 
\end{align}
and for the right ventricle, ``lv'' is substituted with ``rv''.
This condition has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this condition is imposed on the current configuration, and
to utilize the Lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lVendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{epi}}  J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{epi}}$ is usually set
to zero.

@@

Another approach which is used in this thesis is to impose a Robin
type boundary condition at the base
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase, 
\end{align}
or at the epicardium to mimic the pericardium
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvepi.
\end{align}
Here $k$ can be seen as the stiffness of a spring that limit the
movement. The limiting cases, $k = 0$ and $k \rightarrow
\infty$ represent free and fixed boundary respectively.
More complex boundary conditions to mimic the pericardium is also
possible \cite{fritz2014simulation}, but not considered in this thesis.
An overview of the location of the different boundaries for the
bi-ventricular geometry is illustrated in Figure \ref{fig:boundaries}.

@@

\subsection{Force-balance equation}
We will now collect all the terms that makes up to force balance for
the cardiac mechanics problem. Considering the myocardium as an incompressible,
hyperelastic material we obtain the following strong form in the
Lagrangian formulation 
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derive weak
variational form of this equation.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

refer to this approach as the \emph{principle of virtual work}, with states

@@

We have already covered the theory of hyperelasticity and incompressibility in Section
\ref{sec:hyperelasticity} and \ref{sec:incompressibility} respectively
which are types of constitutive relations. In this section we will
cover constitutive relations which only apply to soft living tissue
such as the myocardium.

A complete constitutive model of the mechanical behavior of the
myocardium must account for both the passive and the active response
of the myocardium.

@@

One feature that separate the myocardium from other hyperelastic
materials such as rubber, is its ability to actively generate force
without external loads. This active component of the model can be
incorporated using two fundamentally different approaches known as the
\emph{active stress} and \emph{active strain} formulation.

@@

% Other types of active models also exists \cite{goktepe2014generalized}
% but will not be discssed f

@@

% A word about FEniCS and how we can implement this
% Newtons method, good initial guess, increment pressure.
% Relatively small system, dense matrix, plot stiffness matrix, Direct
% solver.
% FEniCS compilation stored in cache.
% Fibers at quadrature points. Symbolic language.

>>>>

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F = \nabla_{\Xvec} \varphi= \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}
Figure \ref{fig:motion} shows how a point in the reference
configuration is transformed to a point in the current configuration. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.6\textwidth]{chapters/introduction/figures/motion.pdf}
\caption{Relations between the current and reference configuration.}
\label{fig:motion}
\end{figure}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

@@

\subsection{Balance laws and transformations}
In this section we will cover some basic transformations used to
derive the fore-balance equations for the mechanics of the heart.

@@

\subsubsection{Conservation of linear momentum}
Newton's seconds law states that the change in linear momentum equals
the net impulse acting on it. For a continuum material with constant
mass density $\rho$ this implies that
\begin{align}
  \int_{\omega} \rho \dot{\mathbf{v}} \mathrm{d}v = \mathbf{f},
  && \mathbf{f} = \int_{\partial \omega} \mathbf{t} \mathrm{d}s
     + \int_{\omega} \mathbf{b} \mathrm{d}v,
     \label{eq:cons_lin_mom}
\end{align}
where $\mathbf{v}$ is the spatial velocity field, $\mathbf{t}$ is
the traction acting on the boundary, and $\mathbf{b}$ is the body
force. From \emph{Cauchy's stress theorem} we know that there exists a
second order tensor $\sigma$, known as the Cauchy stress tensor that is
related to the traction vector by $\mathbf{t} = \sigma \mathbf{n}$,
where $\mathbf{n}$ is the unit normal in the current configuration.
Using the divergence theorem we get
\begin{align*}
  \int_{\partial \omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom} we arrive at
Cauchy's momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) can be considered negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

\subsubsection{Conservation of angular momentum}
Just like linear momentum, the angular momentum is also a conserved
quantity. We will not go through the derivation, but state that as a
consequence, the Cauchy stress tensor is symmetric
\begin{align}
  \sigma = \sigma^T.
\end{align}

@@

Even though experimental studies have indicated visco-elastic behavior
of the myocardium \cite{dokos2002shear, gultekin2016orthotropic}, a
common assumption is to consider quasi-static behavior, meaning that
the inertial term in \eqref{eq:chauch_momentum_eq} is negligible and
static equilibrium is achieved at all points in the cardiac cycle. Therefore
it is also possible to model the myocardium as a hyperelastic
material,which is a type of elastic material. 
This means that we postulate the existence of
a strain-energy density function $\Psi:\mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$, and that stress is given by the relation
\begin{align}
\FPK = \frac{\partial \Psi(\F)}{\partial \F}.
\end{align}
Since stress has unit Pa, we see that the strain-energy density
function is defined as energy per unit reference volume, and has units
$\frac{\text{Joule}}{m^3}$. 
The strain-energy density function relates  the amount of
energy that is stored within the material in response to a given
strain. Hence, the stresses in a hyperelastic material with a given
strain-energy density function, depend only on the strain, and not the
path for which the material deforms. On the contrary, if the model had
been visco-elastic we would expect to see hysteresis in the
stress/strain curve, but this is not possible for a hyperelastic
material.

\begin{remark}
  The second law of thermodynamics states that the total entropy
  production in a thermodynamic process can never be negative. Elastic
  materials define a special class of materials in which the entropy
  production is zero. Within this thermodynamic framework the
  strain-energy density function coincides (up to a constant) with the
  Helmholtz free energy density.
\end{remark}

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is, the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T = \mathbf{a}_0\otimes\mathbf{a}_0 \},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.

@@

However, from a physical point of view this requirement is too strict
\cite{ball1976convexity}. A slightly weaker requirement is the strong
ellipticity condition which states that \eqref{eq:response_posdev}
should hold for any $\mathbf{H}$ of rank-one, and is analogous to
say that the strain energy function is rank-one convex.

@@

The myocardium contains small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts, this
perfused blood is squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. A material that change its
volume in response to applied loads are referred to as compressible. When the 
volume is preserved we refer to the material as incompressible.
Since 2-4\% is very little, a common assumption in cardiac mechanical modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompressible. The reason for this choice is
purely numerical.

For an incompressible material, only isochoric motions are
possible. This means that the volume of the material does not change during
any deformation, and hence we have the constraint 
\begin{align}
  J = \det(\F) = 1.
  \label{eq:incompressible_cons}
\end{align}
The constraint \eqref{eq:incompressible_cons} can be imposed by
considering the modified strain energy function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + J p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = J^{-1} \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

@@

\subsubsection{Uncoupling of volumetric and isochoric response}
The total strain energy function in \eqref{eq:incomp_strain_energy}
can be written as a sum of isochoric and volumetric components. Let
\begin{align}
  \F =  \F_{\mathrm{vol}} \F_{\mathrm{iso}},
\end{align}
then $ \F_{\mathrm{vol}} =
J^{1/3}\I$ and $\F_{\mathrm{iso}} = J^{-1/3}\F$. For
compressible materials (i.e with $J \neq 1$) it is important to consider
only deviatoric strains in the strain-energy density function, so that
$\Psi = \Psi_{\mathrm{iso}}(\F_{\mathrm{iso}}) +
\Psi_{\mathrm{vol}}(J)$. For incompressible material ($J = 1$), we
have $\F_{\mathrm{vol}} = \I$ so that such a decomposition seems
unnecessary. However, a similar decomposition has shown to be
numerically beneficial \cite{weiss1996finite}. Note that, in this case, a similar
decoupling of the stress tensors has to be done.

@@

Choosing the correct boundary conditions for the model is essential,
and the choice should mimic what is observed in reality. To
physiologically constrain the ventricle in a correct way is difficult,
and different approaches has been proposed.
The boundary condition at the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure. For the left ventricle we have
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lVendoCur, 
\end{align}
and for the right ventricle, ``lv'' is substituted with ``rv''.
This condition has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this condition is imposed on the current configuration, and
to utilize the Lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lVendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{epi}}  J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{epi}}$ is often set to zero as a
simplification.

@@

Another approach which is used in this thesis is to impose a Robin
type boundary condition at the base
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase, 
\end{align}
or at the epicardium to mimic the pericardium
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvepi.
\end{align}
Here $k$ can be seen as the stiffness of a spring that limits the
movement. The limiting cases, $k = 0$ and $k \rightarrow
\infty$ represent free and fixed boundary respectively.
More complex boundary conditions to mimic the pericardium are also
possible \cite{fritz2014simulation}, but not considered in this thesis.
An overview of the location of the different boundaries for the
bi-ventricular geometry is illustrated in Figure \ref{fig:boundaries}.

@@

\subsection{Force-balance equation}
We will now collect all the terms that are involved in the force balance for
the cardiac mechanics problem. Considering the myocardium as an incompressible,
hyperelastic material we obtain the following strong form in the
Lagrangian formulation 
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derive weak
variational form of this equation.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

refer to this approach as the \emph{principle of virtual work}, which states

@@

We have already covered the theory of hyperelasticity and incompressibility in Section
\ref{sec:hyperelasticity} and \ref{sec:incompressibility} respectively
which are types of constitutive relations. In this section we will
cover constitutive relations which only apply to soft living tissue
such as the myocardium. In particular, we will consider a complete
constitutive model of the mechanical behavior of the myocardium that
accounts for both the passive and the active response of the myocardium.

@@

One feature that separates the myocardium from other hyperelastic
materials such as rubber, is its ability to actively generate force
without external loads. This active component of the model can be
incorporated using two fundamentally different approaches known as the
\emph{active stress} and \emph{active strain} formulation.

@@

For a more extensive comparison of the active stress and active strain
approach we refer to \cite{ambrosi2012active,giantesio2017comparison},
and for an overview of other methods to model the active contraction
we refer to \cite{goriely2017five}.

@@
